---
title: "Biostat 203B Homework 2"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Yufan Gong (UID: 305301666)"
format:
  html:
    theme: cosmo
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
knitr:
  opts_chunk: 
    cache: false    
    echo: true
    fig.align: 'center'
    fig.width: 6
    fig.height: 4
    message: FALSE
---

Display machine information for reproducibility:
```{r}
#| eval: false
sessionInfo()
```

Load necessary libraries (you can add more as needed).
```{r setup, message=FALSE}
library(data.table)
library(lubridate)
library(R.utils)
library(tidyverse)
```

MIMIC data location
```{r}
mimic_path <- "~/mimic"
```

In this exercise, we use tidyverse (ggplot2, dplyr, etc) to explore the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2023winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

```{bash eval=FALSE}
for folder in core hosp icu MIMIC-reduce-chartevents
  do 
    for datafile in /Volumes/Data/mimic-iv-1.0/$folder
      do 
        ln -s $datafile ~/mimic/$folder
      done
  done
```
It seems like R requires to use the full path pointing to the original file.
Here I re-link each file within each folder.

Display the contents of MIMIC data folder. 
```{r}
c("/", "/core", "/hosp", "/icu") %>% 
  map(~str_c("ls -l ", mimic_path, .x) %>% 
        system(intern = TRUE)
      )

```

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package. 

Which function is fastest? Is there difference in the (default) parsed data types? (Hint: R function `system.time` measures run times.)

For later questions, we stick to the `read_csv` in tidyverse.

```{r, message=FALSE}
list(read.csv, read_csv, fread) %>% 
  map(
    ~str_c(mimic_path, "/core", "/admissions.csv.gz") %>% 
      . %>% 
      system.time
    ) 

list(read.csv, read_csv, fread) %>% 
  map(
    ~str_c(mimic_path, "/core", "/admissions.csv.gz") %>% 
      . %>% 
      map(class)
    ) 
  
```
It seems like `fread` is the fastest way to read a compressed csv file.
The default data types are different between the three methods. In `read.csv` and `fread`, all numeric variables were classified as integer, while they remained numeric in `read_csv`. All date variables were classified as character in `read.csv`, while they remained as date format in `read_csv` and `fread`.

## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/icustays.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `icustatys.csv.gz` as a tibble `icustays_tble`. 

```{r, message=FALSE}
icustays_tble <- str_c(mimic_path, "/icu/icustays.csv.gz") %>% 
  read_csv()
```


2. How many unique `subject_id`? Can a `subject_id` have multiple ICU stays? 
```{r}
icustays_tble %>% 
  select(subject_id) %>% 
  distinct() %>% 
  nrow()
```
There are 53150	unique `subject_id`. Yes, one `subject_id` can have multiple ICU stays.

3. Summarize the number of ICU stays per `subject_id` by graphs. 

```{r}
icustays_tble %>% 
  group_by(subject_id) %>% 
  summarise(count = n(), .groups = "keep") %>% 
  ggplot(aes(x = count)) +
  geom_bar() +
  labs(
    x = "Number of ICU stays",
    y = "Number of subject ids"
  ) +
  theme_classic()
```


4. For each `subject_id`, let's only keep the first ICU stay in the tibble `icustays_tble`. (Hint: `slice_min` and `slice_max` may take long. Think alternative ways to achieve the same function.)

```{r}
icustays_tble_first <- icustays_tble %>% 
  group_by(subject_id) %>% 
  arrange(intime) %>% 
  summarise(first_stay = first(intime))
```


## Q3. `admission` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/admissions/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/core/admissions.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `admissions.csv.gz` as a tibble `admissions_tble`.

```{r, message=FALSE}
admissions_tble <- str_c(mimic_path, "/core/admissions.csv.gz") %>% 
  read_csv()
```


2. Let's only keep the admissions that have a match in `icustays_tble` according to `subject_id` and `hadm_id`.

```{r}
admissions_tble %>% 
  semi_join(icustays_tble, by = c("subject_id", "hadm_id"))
```


3. Summarize the following variables by graphics. 

    - admission year  
    - admission month  
    - admission month day  
    - admission week day  
    - admission hour (anything unusual?)  
    - admission minute (anything unusual?)  
    - length of hospital stay (anything unusual?)   
    
```{r}
admissions_tble %>% 
  semi_join(icustays_tble, by = c("subject_id", "hadm_id")) %>% 
  mutate(year = str_sub(as.character(admittime), start = 1, end = 4)) %>% 
  ggplot(aes(x = year)) +
  geom_bar() +
  labs(
    x = "Year",
    y = "Number of subject ids"
  ) +
  coord_flip() +
  theme_classic()
```

    
## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/patients/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/core/patients.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/core/patients/>) as a tibble `patients_tble` and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).

```{r, message=FALSE}
patients_tble <- str_c(mimic_path, "/core/patients.csv.gz") %>% 
  read_csv() %>% 
  semi_join(icustays_tble, by = "subject_id")
```


2. Summarize variables `gender` and `anchor_age`, and explain any patterns you see.

```{r}
patients_tble %>% 
  group_by(gender) %>% 
  ggplot(aes(x = anchor_age)) + 
  geom_bar() + 
  facet_wrap(~gender)
```
```{r}
patients_tble %>% 
  ggplot(aes(x = anchor_age)) +
  geom_bar()
```

## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/hosp/labevents.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

`d_labitems.csv.gz` is the dictionary of lab measurements. 
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/hosp/d_labitems.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Find how many rows are in `labevents.csv.gz`.

```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/hosp/labevents.csv.gz"), 
    " | wc -l"
    ), 
  intern = TRUE
)
```


2. We are interested in the lab measurements of creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), and glucose (50931). Retrieve a subset of `labevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `labevents_tble`. 

    Hint: `labevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` option in the `read_csv` function to reduce the memory burden. It took my computer 5-10 minutes to ingest this file. If your computer really has trouble importing `labevents.csv.gz`, you can import from the reduced data file `labevents_filtered_itemid.csv.gz`.
    
```{r}
labevents_tble <- str_c(mimic_path, "/hosp/labevents.csv.gz") %>% 
  read_csv(col_select()) %>% 
  filter(item_id %in% c("50971", "50983", "50902", "50882", 
                        "51221", "51301", "50931"))
```


3. Further restrict `labevents_tble` to the first lab measurement during the ICU stay. 

4. Summarize the lab measurements by appropriate numerics and graphics. 

## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/chartevents.csv.gz"), 
    " | head"), 
  intern = TRUE
)
```
`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`. 
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/d_items.csv.gz"), 
    " | head"), 
  intern = TRUE
)
```

1. We are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `chartevents_tble`.

    Hint: `chartevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` option in the `read_csv` function to reduce the memory burden. It took my computer >15 minutes to ingest this file. If your computer really has trouble importing `labevents.csv.gz`, you can import from the reduced data file `chartevents_filtered_itemid.csv.gz`.

2. Further restrict `chartevents_tble` to the first vital measurement during the ICU stay. 

3. Summarize these vital measurements by appropriate numerics and graphics. 

## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are the first ICU stay of each unique adult (age at admission > 18) and columns contain at least following variables  

- all variables in `icustays.csv.gz`  
- all variables in `admission.csv.gz`  
- all variables in `patients.csv.gz`  
- first lab measurements during ICU stay  
- first vital measurements during ICU stay
- an indicator variable `thirty_day_mort` whether the patient died within 30 days of hospital admission (30 day mortality)

## Q8. Exploratory data analysis (EDA)

Summarize following information using appropriate numerics or graphs.

- `thirty_day_mort` vs demographic variables (ethnicity, language, insurance, marital_status, gender, age at hospital admission)

- `thirty_day_mort` vs first lab measurements

- `thirty_day_mort` vs first vital measurements

- `thirty_day_mort` vs first ICU unit
